Comparing \pkg{binsegRcpp} with other implementations of binary segmentation for change-point detection

### 2025-05-06 article.Rnw
figure-heap-refs-other

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Toby Dylan Hocking~\orcid{0000-0002-3146-0865}\\Northern Arizona University}
\Plainauthor{Toby Dylan Hocking}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Comparing binsegRcpp with other implementations of binary segmentation for changepoint detection}
\Plaintitle{Comparing binsegRcpp with other implementations of binary segmentation for changepoint detection}
\Shorttitle{binsegRcpp for changepoint detection}

%% - \Abstract{} almost as usual
\Abstract{
  Binary segmentation is a classic algorithm for detecting
  changepoints in sequential data. In theory, using a simple loss
  function like Gaussian or Poisson, binary segmentation should be
  extremely fast for $N$ data and $K$ segments, $O(N K)$ in the worst
  case, and $O(N \log K)$ in the best case. In practice, existing
  implementations such as \pkg{ruptures} (\proglang{Python} module) and
  \pkg{changepoint} (\proglang{R} package) are substantially slower.
  Additionally, \pkg{changepoint} sometimes returns incorrect results.
  We present the \proglang{R} package \pkg{binsegRcpp}, which provides an efficient
  and correct C++ implementation of binary segmentation.
  We discuss some C++ coding techniques
  which were used to avoid repetition, for efficiency, and for
  readability.  We additionally include detailed theoretical and
  empirical comparisons with other implementations of binary
  segmentation in other packages: \pkg{ruptures}, \pkg{fpop}, \pkg{changepoint} and
  \pkg{wbs}.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{JSS, style guide, comma-separated, not capitalized, \proglang{R}}
\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}

@@

@@

  \section{More technical details} \label{app:technical}
  TODO

>>>>

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Toby Dylan Hocking~\orcid{0000-0002-3146-0865}\\Université de Sherbrooke}
\Plainauthor{Toby Dylan Hocking}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Comparing \pkg{binsegRcpp} with other implementations of binary segmentation for change-point detection}
\Plaintitle{Comparing binsegRcpp with other implementations of binary segmentation for change-point detection}
\Shorttitle{binsegRcpp for efficient change-point detection}

%% - \Abstract{} almost as usual
\Abstract{
  Binary segmentation is a classic algorithm for detecting
  change-points in sequential data. In theory, using a simple loss
  function like Normal or Poisson, binary segmentation should be
  extremely fast for $N$ data and $K$ segments: asymptotically $O(N K)$ time in the worst
  case, and $O(N \log K)$ time in the best case.
  In practice,
  other implementations can be asymptotically slower, and can 
  return incorrect results. We propose \pkg{binsegRcpp}, an \proglang{R} package which
  provides a correct \proglang{C++} implementation, with the expected asymptotic time complexity.
  %% In practice, existing
  %% implementations such as \pkg{ruptures} (\proglang{Python} module) and
  %% \pkg{changepoint} (\proglang{R} package) are asymptotically slower.
  %% Additionally, \pkg{changepoint} sometimes returns incorrect results.
  %% We present the \proglang{R} package \pkg{binsegRcpp}, which provides an efficient
  %% and correct C++ implementation of binary segmentation.
  We discuss several important \proglang{C++} coding techniques,
  and include detailed comparisons with other implementations of binary
  segmentation: \pkg{ruptures}, \pkg{fpop}, \pkg{changepoint}, \pkg{blockcpd}, and
  \pkg{wbs}.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{\proglang{C++}, \proglang{R}, binary segmentation, change-point}
\Plainkeywords{C++, R, binary segmentation, change-point}

@@

TODO \pkg{blockcpd} package TODO \citep{blockcpd2021,Prates2021} .

@@

  \section{More technical details} \label{app:technical}

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth]{figure-heap-pred}
  \caption{\label{fig:heap-pred} Comparing computation time of \pkg{binsegRcpp} with three containers, and two baselines: \pkg{changepoint} and \pkg{ruptures}.}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth]{figure-heap-refs-binsegRcpp}
  \caption{\label{fig:heap-refs} Estimation of asymptotic time (seconds) and memory (kilobytes) usage for \pkg{binsegRcpp}.
    Black curves are empirical measurements, and violet curves are references.
    These data suggest that both list and multiset containers are $O(N)$;
    list is $O(N^2)$ time and multiset is $O(N\log N)$ time.  }
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth]{figure-heap-refs-other}
  \caption{\label{fig:heap-refs} Estimation of asymptotic time (seconds) and memory (kilobytes) usage for \pkg{changepoint} and \pkg{ruptures}.
    Black curves are empirical measurements, and violet curves are references.
    These data suggest that \pkg{changepoint} is $O(N^2)$ space and $O(N^3)$ time;
    \pkg{ruptures} is $O(N)$ space and $O(N^{1.5})$ time.
  }
\end{figure}
